\myTitle

### 2024-03-20 002-GSI_and_Fst.Rmd
Added pop gen summaries

# let's also bootstrap that:
FST_boot <- boot.ppfst(
  dat = hfstat_dat,
  nboot = 1000,
  quant = c(0.025,0.975),
  diploid = TRUE
)

@@

>>>>

@@

# Population Genetic Summaries

We can do all of our standard population genetic summaries here using `full_base2`.

It is pretty standard, but I think that i am going to standardize sample sizes.

```{r}
fbl_no_miss <- full_base2 %>%
  select(-starts_with("NC_037130.1")) %>%  # drop the LFARs
  pivot_longer(
    cols = c(-(indiv:sample_type)),
    names_to = c("locus", "gene_copy"),
    names_pattern = "^(.*)_([12])$"
  ) %>%
  group_by(indiv, locus) %>%
  mutate(isHz = value[1] != value[2]) %>%
  ungroup() %>%
  filter(!is.na(value) & !is.na(isHz))

# now we just summarize by collection and locus
simple_summaries <- fbl_no_miss %>%
  group_by(repunit, collection, locus) %>%
  summarise(
    num_diploids = n() / 2.0,
    obs_Hz = mean(isHz),
    num_alleles = n_distinct(value)
  ) %>%
  ungroup()

# here, we get the minimum number of diploids at each locus
min_gene_copies <- simple_summaries %>%
  group_by(locus) %>%
  summarise(min_gc = min(num_diploids) * 2)

# here is a function to take N samples of size s from a vector
# and return the mean number of distinct elements in it:
subsample_num_alleles <- function(v, s, N=1000) {
  samples <- lapply(
    1:N, 
    function(z) {
      sample(x = v, size = s, replace = FALSE) %>% n_distinct()
    }) %>% unlist()

    # now, return a tibble with columns mean_num_alle_subsamp an fract_polymorphic_subsamp
    tibble(
      mean_num_alle_subsamp = mean(samples),
      fract_polymorphic_subsamp = mean(samples > 1)
    )
}

# we will now use that min_diploid number to downsample
# each collection randomly, without replacement 1000 times and 
# calculate the mean number of alleles.
sub_sampled_locus_num_alle <- fbl_no_miss %>%
  left_join(min_gene_copies, by = join_by(locus)) %>%
  group_by(repunit, collection, locus) %>%
  summarise(
    mean_num_alleles = list(subsample_num_alleles(value, min_gc[1], N = 100))
  ) %>%
  ungroup() %>%
  unnest(mean_num_alleles)

# and now we can calculate the means over loci of those
subby <- sub_sampled_locus_num_alle %>%
  group_by(repunit, collection) %>%
  summarise(
    mean_num_alle = mean(mean_num_alle_subsamp),
    mean_fract_poly = mean(fract_polymorphic_subsamp)
  )

# here we will get the estimated allele freqs and, from those, calculate
# expected heterozygosity as 1 - \sum_alleles freq^2
expHz <- fbl_no_miss %>%
  group_by(repunit, collection, locus, value) %>%
  summarise(
    n = n()
  ) %>%
  mutate(freq = n / sum(n)) %>%
  group_by(repunit, collection, locus) %>%
  summarise(exp_Hz = 1 - sum(freq ^2)) %>%
  ungroup()

# now, join that on there:
locus_summaries <- simple_summaries %>%
  left_join(expHz, by = join_by(repunit, collection, locus))

# and now we can summarise there into means and SDs
mean_Hzs_etc <- locus_summaries %>%
  group_by(repunit, collection) %>%
  summarise(
    mean_num_dip = mean(num_diploids),
    mean_expHz = mean(exp_Hz),
    mean_obsHz = mean(obs_Hz),
  )

# OK, join the subsampled numbers onto those
tibble_for_paper <- subby %>%
  left_join(mean_Hzs_etc, by = join_by(repunit, collection)) %>%
  mutate(collection = factor(collection, levels = unique(pop_labels$collection))) %>%
  arrange(collection) %>%
  ungroup() %>%
  select(collection, mean_num_dip, mean_num_alle, mean_fract_poly, mean_expHz, mean_obsHz) %>%
  mutate(
    collection = as.character(collection),
    mean_num_dip = sprintf("%.1f", mean_num_dip),
    mean_num_alle = sprintf("%.2f", mean_num_alle),
    mean_fract_poly = sprintf("%.2f", mean_fract_poly),
    mean_expHz = sprintf("%.3f", mean_expHz),
    mean_obsHz = sprintf("%.3f", mean_obsHz)
  ) %>%
  rename(
    Code = collection,
    `$\\bar{N}$` = mean_num_dip,
    `$\\bar{N}_\\mathrm{A,ss}$` = mean_num_alle,
    `$\\bar{P}_\\mathrm{poly,ss}$` = mean_fract_poly,
    `$\\bar{\\mathrm{H}}_\\mathrm{exp}$` = mean_expHz,
    `$\\bar{\\mathrm{H}}_\\mathrm{obs}$` = mean_obsHz,
  )

tibble_for_paper$Code[1] <- sprintf("\\hline %s", tibble_for_paper$Code[1])

# then write that table out:
write_delim(
  tibble_for_paper,
  delim = "&",
  file = "tex/inputs/popgen-summary.tex", 
  eol = "\\tabularnewline\n", 
  quote = "none"
)
```

# Finally, get the histogram of the total number of alleles per locus

```{r}
numa <- fbl_no_miss %>%
  group_by(locus) %>%
  summarise(
    num_alle = n_distinct(value)
  ) %>%
  count(num_alle) %>%
  bind_rows(tibble(num_alle = 9, n = 0))

gbp <- ggplot(numa, aes(x = factor(num_alle), y = n)) +
  geom_col(fill = "blue") +
  theme_bw() +
  xlab("Total number of alleles") +
  ylab("Number of loci")

ggsave(gbp, filename = "tex/images/num-alle-barplot.pdf", width = 4, height = 3)
```

### 2024-03-20 tex/body-text-of-paper.tex
Added pop gen summaries

(We need to talk about how these were split into spring run, fall-run, winter 
run, etc.
In total, the samples in the baseline represent fish from five different Evolutionarily Significant 
Units (ESUs) of Chinook salmon \citep{waples1991pacific} A summary of the different collections 
%%%%%%%%%%%%%%%%
\begin{table*}
\caption{\footnotesize Collections of fish in the reference baseline.  The Sampling Months column shows
a pictorial representation of the proportion of each collection sampled in each month.  Colors in
this column correspond to run-timing groups as described in the caption of Fig.~\protect\ref{fig:map}.}
{\small 
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} llllllrcr}
\hline\hline\vspace*{0.4ex}
\input{inputs/samples-table.tex}
\end{tabular*}
}
\rule{3cm}{0.3pt}

@@

Analysis of the mapping of amplicons for Chr34:828,768 
showed that many ($\approx$89\%) of the reads were off-target (aligning to other chromosomes, 
etc.), making it costly (in terms of sequencing effort) to include the marker in the baseline panel. 
Consequently it was dropped.  The frequency of the different alleles at the two remaining loci
within different reporting units in the baseline are shown in Table~\ref{tab:lfar-freqs}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Allele frequencies across reporting units of the two late-fall associated
markers.  Frequencies are given for the alleles most common amongst late-fall Chinook: nucleotide 
{\tt G} at locus Chr34:865,057 and {\tt A} at Chr34:1,063,084.  $N$ is the total number of gene 
copies available for estimating the allele frequency.}
\label{tab:lfar-freqs}
{\small
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}} lrrrr}
\hline\hline
& \multicolumn{2}{c}{\underline{Chr34:865,057}} & \multicolumn{2}{c}{\underline{Chr34:1,063,084}} \\
Reporting Unit & freq. & $N$ & freq. & $N$ \\ \hline
\input{inputs/lfar-freqs.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule\vspace*{0.3ex}\hrule
\end{table}
%%%%%%%%%%%%%%%
\subsection*{Population genetic analyses}

@@

The results are summarized in Fig.~\ref{fig:fprs}.
%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
\newcommand{\fprcap}{\footnotesize False positive rates as a function of false negative
rates for distinguishing between different relationships.  {\bf (a)} False positive rates calculated
using importance sampling of unrelated pairs of individuals being incorrectly categorized as parent-offspring (PO),
full-sibling (FS) or half-sibling (HS) pairs.  {\bf (b)}  False positive rates calculated using regular Monte Carlo
sampling of avuncular pairs (AN: aunt-niece, uncle-nephew, etc.) or full-siblings (FS) or half-siblings (HS) being 
incorrectly categorized as parent-offspring (PO); as well as FPRs of parent-offspring (PO) pairs or half-sibling (HS) 
pairs being incorrectly inferred as full siblings (FS).  In both {\em a} and {\em b}, 100,000 Monte Carlo replicates 
were simulated.  Importance sampling allows estimation of very small false positive rates for unrelated pairs, but 
cannot be applied to truly related pairs.  Consequently, it is difficult to accurately estimate small probabilities 
($10^{-4}$ or less) in {\bf b}, as indicated by the error bars, which represent the estimate $\pm 2 \times s$ where $s$ is the estimated standard error of the mean of the Monte Carlo sample.  }
\begin{center}
\includegraphics[width=\textwidth]{images/fpr-fnr-figure-crop.pdf}
\end{center}
\caption[\fprcap]{\fprcap}
\label{fig:fprs}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%

>>>>

(We need to talk about how these were split into spring run, fall-run, winter 
run, etc.
In total, the samples in the baseline represent fish from five different Evolutionarily Significant 
Units (ESUs) of Chinook salmon \citep{waples1991pacific} A summary of the different collections 
%%%%%%%%%%%%%%%%
\begin{table*}
\caption{\footnotesize Collections of fish in the reference baseline.  The Sampling Months column shows
a pictorial representation of the proportion of each collection sampled in each month.  Colors in
this column correspond to run-timing groups as described in the caption of Fig.~\protect\ref{fig:map}.}
\label{tab:samples}
{\small 
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} llllllrcr}
\hline\hline\vspace*{0.4ex}
\input{inputs/samples-table.tex}
\end{tabular*}
}
\rule{3cm}{0.3pt}

@@

Analysis of the mapping of amplicons for Chr34:828,768 
showed that many ($\approx$89\%) of the reads were off-target (aligning to other chromosomes, 
etc.), making it costly (in terms of sequencing effort) to include the marker in the baseline panel. 
Consequently it was dropped.  The frequency of the different alleles at the two remaining loci
within different reporting units in the baseline are shown in Table~\ref{tab:lfar-freqs}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Allele frequencies across reporting units of the two late-fall associated
markers.  Frequencies are given for the alleles most common amongst late-fall Chinook: nucleotide 
{\tt G} at locus Chr34:865,057 and {\tt A} at Chr34:1,063,084.  $N$ is the total number of gene 
copies available for estimating the allele frequency.}
\label{tab:lfar-freqs}
{\small
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}} lrrrr}
\hline\hline
& \multicolumn{2}{c}{\underline{Chr34:865,057}} & \multicolumn{2}{c}{\underline{Chr34:1,063,084}} \\
Reporting Unit & freq. & $N$ & freq. & $N$ \\ \hline
\input{inputs/lfar-freqs.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule\vspace*{0.3ex}\hrule
\end{table}
%%%%%%%%%%%%%%%
\subsection*{Population genetic summaries}

Barplot of the total number of alleles per locus in the whole data set
is Figure~\ref{fig:num-alle}.
%%%%%%%%%%%%%%%%%%%%%% 
\begin{figure}
\newcommand{\numacap}{\footnotesize Number of loci with different
total numbers of alleles in the data set.}
\begin{center}
\includegraphics[width=\columnwidth]{images/num-alle-barplot.pdf}
\end{center}
\caption[\numacap]{\numacap}
\label{fig:num-alle}
\end{figure}
%%%%%%%%%%%%%%%%%%%%

And the table of the popgen summary is Table~\ref{tab:pg-summ}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Simple population genetic summaries. All quantities are means
over loci.  $\bar{N}$ is the number of fish with observed genotypes; $\bar{N}_\mathrm{A,ss}$ is the
average number of alleles and $\bar{P}_\mathrm{poly,ss}$ the fraction of polymorphic loci,
after sub-sampling to the smallest sample size per locus; $\bar{H}_\mathrm{exp}$ and
$\bar{H}_\mathrm{obs}$ are expected and observed heterozygosity, respectively. Population
codes are as given in Table~\ref{tab:samples}}
\label{tab:pg-summ}
{\small
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}} lrrrrr}
\hline\hline
\input{inputs/popgen-summary.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule\vspace*{0.3ex}\hrule
\end{table}
%%%%%%%%%%%%%%%

@@

The results are summarized in Fig.~\ref{fig:fprs}.
%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
\newcommand{\fprcap}{\footnotesize False positive rates as a function of false negative
rates for distinguishing between different relationships.  {\bf (a)} False positive rates calculated
using importance sampling of unrelated pairs of individuals being incorrectly categorized as parent-offspring (PO),
full-sibling (FS) or half-sibling (HS) pairs.  {\bf (b)}  False positive rates calculated using regular Monte Carlo
sampling of avuncular pairs (AN: aunt-niece, uncle-nephew, etc.) or full-siblings (FS) or half-siblings (HS) being 
incorrectly categorized as parent-offspring (PO); as well as FPRs of parent-offspring (PO) pairs or half-sibling (HS) 
pairs being incorrectly inferred as full siblings (FS).  In both {\em a} and {\em b}, 100,000 Monte Carlo replicates 
were simulated.  Importance sampling allows estimation of very small false positive rates for unrelated pairs, but 
cannot be applied to truly related pairs.  Consequently, it is difficult to accurately estimate small probabilities 
($10^{-4}$ or less) in {\bf b}, as indicated by the error bars, which represent the estimate $\pm 2s$ where $s$ is the estimated standard error of the mean of the Monte Carlo sample.  }
\begin{center}
\includegraphics[width=\textwidth]{images/fpr-fnr-figure-crop.pdf}
\end{center}
\caption[\fprcap]{\fprcap}
\label{fig:fprs}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%
